agdarsec -- Total Parser Combinators

### 2017-12-10 doc/paper/main.tex
[ done ] following the reviewers' comments

Parser combinators have made functional languages such as Haskell
shine. They are a prime example of the advantages Embedded Domain
Specific Languages provide the end user. She not only has access
to a library of powerful and composable abstractions but she is also
able to rely on the host language's existing tooling and libraries.
She also gets feedback from the static analyses built in the compiler
(e.g. type and coverage checking) and can exploit the expressivity
of the host language to write generic parsers thanks to polymorphism
and higher order functions.

However she only gets the guarantees the host language is willing
to give. In non-total programming languages such as Haskell this
means she will not be prevented from writing parsers which will
unexpectedly fail on some (or even all!) inputs. Handling a
left-recursive grammar is perhaps the most iconic pitfall leading
beginners to their doom: a parser never making any progress.

@@

@@

This stanza directly translates to the following Haskell type (we
use a \textbf{newtype} wrapper to have cleaner error messages):

@@

It is naturally possible to run such a parser and try to extract
a value from a successful parse. Opinions may vary on whether a
run with leftover characters can be considered successful or not.
We decide against it. This is not crucial as both styles can be
mutually emulated by either providing an 'end of file' parser
guaranteeing that only runs with no leftovers are successful or
by extending a grammar with a dummy case consuming the rest of
the input string.

@@

@@

However what makes parsers interesting is that they recognize
structure. As such, they need to reject invalid inputs. The
parser only accepting decimal digit is a bare bones example.
It can be implemented in terms of \parser{guard}, a higher
order parser checking that the value returned by its argument
abides by a predicate which can easily be implemented using
functions from the standard library.

@@

These two definitions are only somewhat satisfactory: the
result of the \parser{digit} parser is still \emph{stringly-typed}.
Instead of using a predicate to decide whether to keep the
value, we can opt for a validation function of type
\mbox{\textit{a → Maybe b}} which returns
a witness whenever the check succeeds. To define this
alternative \parser{guardM} we can once more rely on code
already part of the standard library.

In our concrete example of recognizing a digit, we want to
return an \type{Int} corresponding to it. Once more the
standard library has just the right function to use together
with \parser{guardM}.

\shgrab{guard2}
\hgrab{digit2}

@@

% insert alternative instance
% interesting use case? replicateM?

@@

Our first example of a higher order parser was \parser{guardM}
which takes as arguments a validation function as well as another
parser and produces a parser for the type of witnesses returned
by the validation function.

The two parsers \AB{some} and \AB{many} turn a parser for elements
into ones for non-empty and potentially empty lists of such elements
respectively. They concisely showcase the power of mutual recursion,
higher-order functions and the \type{Functor}, \type{Applicative},
and \type{Alternative} structure.

@@

\remark{Non-Commutative} The disjunction combinator is
non-commutative. As such the definitions of \textit{some} and
\textit{many} will try to produce the longest list as possible
as opposed to a flipped version of \text{many} which would start
by returning the empty list and slowly offer longer and longer
matches.

@@

The ability to parse recursive grammars by simply declaring
them in a recursive manner is however dangerous: unlike type
errors which are caught by the typechecker and partial covers
in pattern matchings which are detected by the coverage checker,
termination is not guaranteed here.

@@

The typical solution to this problem is to introduce two 'tiers'
of expressions: the \parser{base} ones which can only be whole
expressions if we consume an opening parenthesis first and the
\parser{expr'} ones which are left-associated chains of \parser{base}
expressions connected by '+'.\label{parser:expr}

@@

This approach can be generalised when defining more complex
languages by having even more tiers, one for each
\emph{fixity level}\label{fixity}, see for instance~\ref{example:expr}.
An extended language of arithmetic expressions would for
instance distinguish the level at which addition and
substraction live from the one at which multiplication and
division do.

Our issue with this solution is twofold. First, although we did
eventually managed to build a parser that worked as expected,
the compiler was unable to warn us and guide us towards this
correct solution. Additionally, the blatant partiality of
some of these definitions means that these combinators and
these types are wholly unsuitable in a total setting. We could, of
course use an escape hatch and implement our parsers in Haskell
but that would both be unsafe and mean we would not be able
to run them at typechecking time which we may want to do if
we embed checked examples in our software's documentation,
or use constant values (e.g. filepaths).

@@

The first kind of combinators corresponds to operations on
sets which are lifted to indexed sets by silently propagating
the index. We only show the ones we will use in this paper:
the pointwise arrow and product types and the constant function.

\sagrab{Relation/Unary/Indexed}{arrow}

\begin{minipage}{0.55\textwidth}
  \agrab{Relation/Unary/Indexed}{product}
\end{minipage}
\begin{minipage}{0.30\textwidth}
\agrab{Relation/Unary/Indexed}{constant}
\end{minipage}

\remark{Mixfix Operators} In Agda underscores correspond to
positions in which arguments are to be inserted. It may be a
bit surprising to see infix notations for functions taking
three arguments but they are only meant to be partially applied.
\medskip{}

The second kind of combinator turns an indexed set into a set
by universally quantifying over the index.

\sagrab{Relation/Unary/Indexed}{forall}

\remark{Implicit Arguments} We use curly braces so that the
index we use is an \emph{implicit} argument we will never
have to write: Agda will fill it in for us by unification.
\medskip{}

We can already see the benefits of using these aliases. For instance
the fairly compact expression \AF{[} (\AF{κ} \AB{P} \AF{⊗} \AB{Q}) \AF{⟶} \AB{R} \AF{]}
corresponds to the more verbose type
\AS{∀} \{\AB{n}\} \AS{→} (\AB{P} \AF{×} \AB{Q} \AB{n}) \AS{→} \AB{R} \AB{n}.

Last but not least, the \BOX{} type constructor takes a \AD{ℕ}-indexed
set and produces the set of valid recursive calls for a function defined
by course-of-values recursion: \BOX{} \AB{A} \AB{n} associates to each
\AB{m} strictly smaller than \AB{n} a value of type \AB{A} \AB{m}.
This construct, analogous to the later modality showing up in Guarded
Type Theory~\cite{vezzosi2015guarded}, empowers the user to give precise
types in a total language to programs commonly written in partial ones
(see e.g. the definition of \textit{fix} below).

@@

\remark{Record Wrapper} Instead of defining \BOX{} as a
function like the other combinators, we wrap the corresponding
function space in a record type. This prevents normalisation from
unfolding the combinator too eagerly and makes types more readable
during interactive development.

\remark{Irrelevance} The argument stating that \AB{m} is strictly
smaller than \AB{n} is preceded by a dot. In Agda, it means that
this value is irrelevant and can be erased by the compiler.
In Coq, we would define the relation \AF{\_<\_} in \type{Prop} to
achieve the same.

This \BOX{} type combinator has some useful value combinators.
The first thing we can notice is the fact that \textbf{\BOX{} is a
functor}; that is to say that given a natural transformation from
\AB{A} to \AB{B}, we can define a natural transformation from
\BOX{} \AB{A} to \BOX{} \AB{B}.

@@

\remark{Copatterns} The definition of \AF{map} uses
the \BOX{} field named \ARF{call} on the \emph{left hand side}.
This is a copattern~\cite{abel2013copatterns}, meaning that we
explain how the definition is \emph{observed} (via \ARF{call})
rather than \emph{constructed} (via \AIC{mkBox}).

Because \AF{\_<\_} is defined in terms of \AD{\_≤\_}, \AF{≤-refl}
which is the proof that \AD{\_≤\_} is reflexive is also a proof
that any \AB{n} is strictly smaller than \AN{1} \AF{+} \AB{n}.
We can use this fact to write the following \AF{extract} function:

@@

As already highlighted in Section~\ref{sec:haskellproblem}, \AB{some}
and \AB{many} can yield diverging computations if the parser they are
given as an argument succeeds on the empty string. To avoid any such
issue, we adopt a radical solution: for a parser's run to be considered
successful, it must have consumed some of its input.

@@

\remark{Implicit Field} Just like the arguments to a function can be
implicit, so can a record's fields. The user can then leave them out
when building a value and they will be filled in by unification.

Coming back to Fritz Ruehr's rhyme, we can define our own \AR{Parser}
type: a parser for things up to size \AB{n} is a function from strings
of length \AB{m} less than \AB{n} to lists of \AR{Success}es of size \AB{m}.

@@

@@

Unsurprisingly \AF{guardM} is still a valid higher-order combinator:
filtering out results which do not agree with a predicate is absolutely
compatible with the consumption constraint we have drawn. To implement
\AF{guardM} we can once more reuse existing library functions such as
\AF{gfilter} which turns a \mbox{\AD{List} \AB{A}} into a \mbox{\AD{List} \AB{B}}
provided a predicate \mbox{\AB{A} \AS{→} \AD{Maybe} \AB{B}}.

@@

We can immediately use all of these newly-defined combinators to give
a safe, total definition of \AF{some} which takes a parser for \AB{A}
and returns a parser for non-empty lists of \AB{A}s. It is defined as
a fixpoint and proceeds as follows: it either combines a head and a
non-empty tail using \AF{$\_::^{+}\_$} or returns a singleton list.

@@

This inefficiency can be fixed by using a notion of conjunction
\AF{\_<\&?>\_} which allows the second parser to fail whilst still
producing a successful run. The parser \mbox{\AF{some} \AB{p}} can then be
defined as the conjunction of \AB{p} and a potentially failing
recursive call. The function \AF{cons} (whose definition we omit here)
either uses \AF{$\_::^{+}\_$} to put the head and the non-empty
tail together or returns a singleton list if the recursive call
failed.

\agrab{Text/Parser/Combinators}{conjunction2}
\agrab{Text/Parser/Combinators}{goodsome}

@@

As the previous code snippet shows, \AF{\_<\&?>\_} is defined in
terms of a more fundamental notion \AF{\_\&?>>=\_} which is a
combinator analogous to a monad's \textit{bind}. On top of running
two parsers sequentially (with the second one being chosen based on
the result obtained by running the first), it allows the second one to fail
and returns both results.

@@

This crucial definition makes it possible to port a lot of the Haskell
definitions where one would use a parser which does not use any of its
input. And this is possible without incurring any additional cost as
the optimised version of \AF{some} showed.

\section{Left Chains}

@@

We naturally want to include a safe variant of this combinator in our
library. However this definition relies on the ability to simply use
\textit{return} in case it's not possible to parse an additional
constructor and argument and that is something we simply don't have
access to. This forces us to find the essence of \parser{rest}, the
auxiliary definition used in \parser{hchainl}: its first argument is
not just a value, it is a \AR{Success} upon which it builds until it
can't anymore and simply returns.

We define \AF{schainl} as the fixpoint of either \AF{rest} (whose
definition we omit here) or the \AR{Success} that was passed as an
input (the function \AF{\_::$^{r}$} appends a single element at the
end of a list).

@@

The type of \AF{rest} may look intimidating but it has this shape because
we build a list of \AR{Success}es at the \emph{same} index as the input
\AR{Success} and \AR{Parser} which will make this combinator compositional
(as opposed to \parser{some} defined in Section~\ref{parser:some}). Ultimately
all of this complexity only shows up in the implementation of our library:
the end user can happily ignore these details.

@@

As we have mentioned when defining \AF{schainl}, the combinator \AF{hchainl}
we have just implemented does not expect fully-defined parsers as arguments.
As a consequence it can be used inside a fixpoint construction. Both the parser
for the constructor and the one for its \AB{B} argument are guarded whilst the
one for the \AB{A} seed is not. This means that trying to define a left-recursive
grammar by immediately using a recursive substructure on the left is now a type
error. But it still possible to have some on the right or after having consumed
at least one character (typically an opening parenthesis, cf. the \type{Expr}
example in Section~\ref{parser:expr}).

\section{Fully Worked-Out Example}

From \AF{hchainl}, one can derive \AF{chainl1} which is not heterogeneous
and uses the same parser for the seed and the constructors' arguments.
This combinator together with the idea of fixity mentioned in Section~\ref{fixity}
is typically used to implement left-recursive grammars. Looking up the
documentation of the \texttt{parsec} library on hackage~\cite{doc:parsec}
we can find a fine example: an extension of our early arithmetic language.

\shgrab{example}

@@

We can notice four minor changes with respect to the Haskell version. Firstly,
the intermediate parsers need to be declared before being used which effectively
reverses the order in which they are spelt out. Secondly, the recursive calls
are now explicit: in the definition of \type{factor}, \AF{expr} is \AF{map}ped
under the \BOX{} to project the recursive call to the \AD{Expr} \AR{Parser} out
of \AR{Language}. Thirdly, we use \AF{hchainl} instead of \AF{chainl1} because
breaking the grammar into three distinct categories leads us to parsing
\emph{h}eterogeneous left chains. Fourthly, we have to insert calls to \AF{box}
to lift \AR{Parser}s into boxed ones whenever the added guarantee that the call
will be guarded is of no use to us.

@@

Secondly, there is no reason to limit ourselves to \type{Char} as the
unit of information to be processed. Near all of our combinators are
fully polymorphic over the kind of tokens they can deal with. To run
the parser, the user will have to declare an instance of the \AR{Tokenizer}
class corresponding to their use case. The library provides a trivial
one for \AD{Char}.

Thirdly, there is no reason to force the user to get back a \emph{list}
of successes: any alternative monad will do. This means in particular that
a user may for instance instrument a parser with a logging ability to be
able to return good error messages, have a (re)configurable grammar using
a \type{Reader} transformer or use a \type{Maybe} type if they want to make
explicit the fact that their grammar is unambiguous.

@@

The reified approach allows him to define a grammar's semantics in terms
of bags of words and prove sound a variant of Brzozowski derivatives~\citeyear{brzozowski1964derivatives}
as well as study the equational theory of parsers. The current
implementation, based on the Brzozowski derivatives, is however of complexity
at least exponential in the size of the input.

Our approach, although not as successful as Danielsson's when it comes
to certification, is however more lightweight. Using only strong induction
on the natural numbers, it is compatible with languages a lot less powerful
than Agda. Indeed there is no need for good support for mixed induction and
coinduction in the host language. And although we do rely on enforcing invariants
at the type-level, one could mimic these in languages with even weaker type
systems by defining an \emph{abstract} \BOX{} and only providing the user with
our set of combinators which is guaranteed to be safe.

>>>>

Parser combinators have made functional languages such as Haskell
shine. They are a prime example of the advantages Embedded Domain
Specific Languages~\cite{hudak1996building} provide the end user. She not only has access
to a set of powerful and composable abstractions but she is also
able to rely on the host language's existing tooling and libraries.
She can get feedback from the static analyses built in the compiler
(e.g. type and coverage checking) and can exploit the expressivity
of the host language to write generic parsers thanks to polymorphism
and higher order functions.

However she only gets the guarantees the host language is willing
to give. In non-total programming languages such as Haskell this
means she will not be prevented from writing parsers which will
unexpectedly fail on some (or even all!) inputs. Handling a
left-recursive grammar is perhaps the most iconic pitfall leading
beginners to their doom: a parser never making any progress.
Other issues one may want guarantees about range from unambiguity
to complexity with respect to the input's size.

@@

\remark{Agda-centric} Although we do use some Agda-specific techniques
in order to have a codebase as idiomatic as possible, we do not expect
the reader to be well-versed in them. We insert remarks similar to this
one throughout the paper to clarify confusing points, and give pointers
to more in-depth explanations to the interested reader.

This work is however not limited to Agda: it can be ported to other
dependently-typed languages and we have already done so for Coq~\cite{Coq:manual}
(\url{https://github.com/gallais/parseque}) and
Idris~\cite{brady2013idris} (\url{https://github.com/gallais/idris-tparsec}).

@@

This stanza translates to the following Haskell type. We use a \textbf{newtype}
wrapper to have cleaner error messages:

@@

It is naturally possible to run such a parser and try to extract
a value from a valid run. Opinions may vary on the exact definition
of a successful parse: should a run with leftover characters or an
ambiguous result be accepted? We decide against both in the following
code snippet but it is not a crucial point.

@@

Once we are equipped with this type of parsers and a function to run
them, we can start providing some examples of parsers and parser combinators.

@@

However what makes parsers interesting is that they recognize
structure. As such, they need to reject invalid inputs. The
parser only accepting decimal digits is a bare bones example.
It can be implemented in terms of \parser{guard}, a higher
order parser checking that the value returned by its argument
abides by a predicate which can easily be implemented using
functions from the standard library.

@@

These two definitions are only somewhat satisfactory: the
result of the \parser{digit} parser is still \emph{stringly-typed}.
Instead of using a predicate to decide whether to keep the
value, we can opt for a validation function of type
\mbox{\textit{a → Maybe b}} which returns a witness whenever the
check succeeds. To define this refined version of \parser{guard}
called \parser{guardM} we can again rely on the standard library:

\shgrab{guard2}

\begin{itemize}
  \item \textit{traverse f} of type \mbox{\textit{(String, a) → Maybe (String, b)}}
    takes apart a pair, applies $f$ to the second component and rebuilds the pair
    \emph{under} the \textit{Maybe} type constructor,
  \item \textit{fmap} applies this function to all the elements in the list obtained
    by running the parser $p$,
  \item and \textit{catMaybes} of type \mbox{\textit{[Maybe (String, b)] → [(String, b)]}}
    only retains the values which successfully passed the test.
\end{itemize}

In our concrete example of recognizing a digit, we want to return the
corresponding \type{Int}. Once more the standard library has just the
right function to use together with \parser{guardM}: \textit{readMaybe}
of (specialised) type \mbox{\textit{String → Maybe Int}}.

\shgrab{digit2}

@@

\textbf{Functor} means that given a function of the right type, we
can alter the values returned by a parser. That is, we have a function
(which corresponds to the infix combinators (<\$>)):

\shgrab{fmap}

\textbf{Applicative} means two things. First, that given a value of type
\type{a}, we can define a parser for values of type \type{a}. Second, that
given a parser for a function and a parser for its argument we can run both
and apply the function to its argument. That is, we have two functions:

\shgrab{applicative}

\textbf{Monad} means that we are entitled to inspect the result of a
first parser to decide which one to run next. This brings us beyond the
realm of context-free grammars. That translates into the existence of
one function:

\shgrab{monad}

\textbf{Alternative} means that for all type \type{a}, \type{Parser a}
forms a monoid. It allows us to take the disjunction of various parsers,
the failure of one leading to the next being used.

\shgrab{alternative}

@@

Our first example of a higher order parser was \parser{guard}
which takes as arguments a validation function as well as another
parser and produces a parser for the type of witnesses returned
by the validation function.

The two parsers \AB{some} and \AB{many} turn a parser for elements
into ones for non-empty and potentially empty lists of such elements
respectively. They concisely showcase the power of mutual recursion,
higher-order functions and the \type{Functor}, \type{Applicative},
and \type{Alternative} structure.
\medskip{}

@@

\remark{Non-Commutative} The disjunction combinator is non-commutative
as ultimately we obtain a list (and not a set) of possible results. As
such the definitions of \textit{some} and \textit{many} will try to
produce the longest list possible as opposed to a flipped version of
\text{many} which would start by returning the empty list and slowly
offer longer and longer matches.

@@

The ability to parse recursive grammars by simply declaring
them in a recursive manner is however dangerous: unlike type
errors which are caught by the typechecker and partial covers
in pattern matchings which are detected by the coverage checker,
termination is not guaranteed.

@@

The typical solution to this problem is to introduce two 'tiers'
of expressions: the \parser{base} ones which can only be whole
expressions if we consume an opening parenthesis first and the
\parser{expr} ones which are left-associated chains of \parser{base}
expressions connected by '+'.\label{parser:expr}

@@

This presentation is still sub-optimal; users would traditionally
be encouraged to used combinators such as \textit{chainl} to avoid
this issue. We will only discuss later in Section~\ref{sec:leftchains}.

This approach can be generalised when defining more complex
languages by having even more tiers, one for each
\emph{precedence level}, see for instance Section~\ref{example:expr}.
An extended language of arithmetic expressions would for
instance distinguish the level at which addition and
subtraction live from the one at which multiplication and
division do.

Our issue with this solution is twofold. First, although we did
eventually manage to build a parser that worked as expected,
the compiler was unable to warn us and guide us towards this
correct solution. Additionally, the blatant partiality of
some of these definitions means that these combinators and
these types are wholly unsuitable in a total setting. We could, of
course use an escape hatch and implement our parsers in Haskell
but that would both be unsafe and mean we would not be able
to run them at typechecking time which we may want to do if
we embed checked examples in our software's documentation,
or use compilation-time configuration via e.g. dependent type
providers~\cite{christiansen2013dependent}.

@@

The first kind of combinators corresponds to operations on
sets which are lifted to indexed sets by silently propagating
the index. We only show the ones we will use in this paper:
the pointwise arrow and product types and the constant function.
The second kind of combinator corresponds to universal quantification:
it turns an indexed set into a set.

\begin{minipage}{0.55\textwidth}
  \agrab{Relation/Unary/Indexed}{arrow}
  \agrab{Relation/Unary/Indexed}{product}
\end{minipage}
\begin{minipage}{0.30\textwidth}
  \agrab{Relation/Unary/Indexed}{constant}
  \agrab{Relation/Unary/Indexed}{forall}
\end{minipage}

\remark{Mixfix Operators} In Agda underscores correspond to
positions in which arguments are to be inserted. It may be a
bit surprising to see infix notations for functions taking
three arguments but they are only meant to be partially applied.
\remark{Implicit Arguments} We use curly braces so that the
index we use is an \emph{implicit} argument we will never
have to write: Agda will fill it in for us by unification.
\medskip{}

We can already see the benefits of these aliases. For instance
the fairly compact expression \AF{[} (\AF{κ} \AB{P} \AF{⊗} \AB{Q}) \AF{⟶} \AB{R} \AF{]}
corresponds to the more verbose type
\AS{∀} \{\AB{n}\} \AS{→} (\AB{P} \AF{×} \AB{Q} \AB{n}) \AS{→} \AB{R} \AB{n}.

Last but not least, we introduce a type constructor which takes a
\AD{ℕ}-indexed set and produces the set of valid recursive calls
for a function defined by course-of-values recursion. By analogy
to modal logic we call it \BOX{} after the ``necessity'' modality
whose interpretation in a Kripke semantics is eerily similar to
our type constructor.

@@

\remark{Record Wrapper} Instead of defining \BOX{} as a
function like the other combinators, we wrap the
function space in a record type. This prevents normalisation from
unfolding the combinator too eagerly and makes types more readable
during interactive development.

\remark{Irrelevance} The argument stating that \AB{m} is strictly
smaller than \AB{n} is preceded by a dot. In Agda, it means that
this value is irrelevant and can be erased by the compiler.
In Coq, we would define the relation \AF{\_<\_} in \type{Prop} to
achieve the same.
\medskip{}

This construct can also be understood as analogous to the later modality
showing up in Guarded Type Theory~\cite{vezzosi2015guarded}. It empowers
the user to give precise types in a total language to programs commonly
written in partial ones (see e.g. the definition of \textit{fix} below).
The first thing we can notice is the fact that \textbf{\BOX{} is a
functor}; that is to say that given a natural transformation from
\AB{A} to \AB{B}, we can define a natural transformation from
\BOX{} \AB{A} to \BOX{} \AB{B}.

@@

\remark{Copatterns} The definition of \AF{map} uses
the \BOX{} field named \ARF{call} on the \emph{left hand side}.
This is a copattern~\cite{abel2013copatterns}, meaning that we
explain how the definition is \emph{observed} (via \ARF{call})
rather than \emph{constructed} (via \AIC{mkBox}).
\medskip{}

Because less than (\AF{\_<\_}) is defined in terms of less than
or equal (\AD{\_≤\_}), \AF{≤-refl} which is the proof that \AD{\_≤\_}
is reflexive is also a proof that any \AB{n} is strictly smaller than
\AN{1} \AF{+} \AB{n}. We can use this fact to write the following
\AF{extract} function:

@@

As already highlighted in Section~\ref{sec:haskellproblem}, \AB{some}
and \AB{many} can yield diverging computations if the parser they are
given as an argument succeeds on the empty string. To avoid any such
issue, we adopt a radical solution: for a parser's run to be considered
successful, it must have consumed some of its input. Some nullability
can be recovered later (see Section~\ref{sec:failure}) when defining
combinators where one of the sub-parses is allowed to fail.

@@

\remark{Implicit Field} Like the arguments to a function can be implicit,
so can a record's fields. The user can leave them out when building a
value: they will be filled in by unification.
\medskip{}

Coming back to Fritz Ruehr's rhyme, we can define our own \AR{Parser}
type: a parser for things up to size \AB{n} is a function from strings
of length \AB{m} less or equal to \AB{n} to lists of \AR{Success}es of size \AB{m}.

@@

\subsection{Our First Combinators}

@@

Unsurprisingly \AF{guardM} is still a valid higher-order combinator:
filtering out results which do not agree with a predicate is absolutely
compatible with the consumption constraint we have drawn. To implement
\AF{guardM} we can once more reuse existing library functions. Relying
this time on Agda's standard library rather than Haskell's, the set of
available function is slightly different. We use for instance \AF{gfilter}
which turns a \mbox{\AD{List} \AB{A}} into a \mbox{\AD{List} \AB{B}}
provided a predicate \mbox{\AB{A} \AS{→} \AD{Maybe} \AB{B}} and combine
\AF{sequence} and \AR{Success}'s \AF{map} to obtain a function akin to
\textit{traverse}.

@@

We can immediately use all of these newly-defined combinators to give
a safe, total definition of \AF{some} which takes a parser for \AB{A}
and returns a parser for \AD{List$^+$} \AB{A}, the type of non-empty
lists of \AB{A}s. It is defined as a fixpoint and proceeds as follows:
it either combines a head and a non-empty tail using \mbox{\AF{$\_::^{+}\_$} :
  \AB{A} $\rightarrow$ \AD{List$^+$} \AB{A} $\rightarrow$
  \AD{List$^+$} \AB{A}} or returns a singleton list.

@@

\subsection{Failure is Sometimes an Option}\label{sec:failure}

This inefficiency can be fixed by introducing the notion of a potentially
failing sub-parse. We use the convention that \AF{?} marks the argument
of a combinator which is allowed to fail e.g. \AF{<\&?>} is the version
of the conjunction \AF{<\&>} whose second argument is allowed to fail
whilst \AF{<?\&>} may let its first argument do so. Which, in terms of
types, translates to:

\agrab{Text/Parser/Combinators}{conjunction2}

The $\textit{some}~\textit{p} \mathbin{<\!\!|\!\!>} \textit{pure}~ [\,]$ pattern
used in the definition of \textit{some} can be translated in our total
setting to a recursive call which is allowed to fail. This leads to the following
rethought definition of \mbox{\AF{some} \AB{p}}. The inefficiency of the previous
version has disappeared: \AB{p} is run once and dependening on the success
or failure of the recursive call it is either added to a non-empty list of
values or returned as a singleton.

\agrab{Text/Parser/Combinators}{goodsome}

@@

The potentially failing conjunction combinator \AF{\_<\&?>\_} can be
generalised to a more fundamental notion \AF{\_\&?>>=\_} which is a
combinator analogous to a monad's \textit{bind}. On top of running
two parsers sequentially (with the second one being chosen based on
the result obtained by running the first), it allows the second one to
fail and returns both results.

@@

These definitions make it possible to port a lot of the Haskell
definitions where one would use a parser which does not use any of its
input. Instead of encoding a potentially failing parse using the pattern
$\textit{p} \mathbin{<\!\!|\!\!>} \textit{pure}~ v$, we can explicitly
use a combinator acknowledging the authorized failure. And this is possible
without incurring any additional cost as the optimised version of \AF{some}
showed.

\subsection{Left Chains}\label{sec:leftchains}

@@

We naturally want to include a safe variant of this combinator in our
library. However this definition relies on the ability to simply use
\textit{pure} in case it's not possible to parse an additional
constructor and argument and that is something we simply don't have
access to.

This forces us to find the essence of \parser{rest}, the auxiliary definition
used in \parser{hchainl}: its first argument is not just a value, it is a
\AR{Success} upon which it builds until it can't anymore and simply returns.
We define \AF{schainl} according to this analysis: it is a bare bones version
of \textit{hchainl}'s \textit{rest} where \textit{con} and \textit{arg} have
already been replaced by a single \textit{con} function.

@@

A key thing to notice is that we build a list of \AR{Success}es at the
\emph{same} index as the input \AR{Success} and \AR{Parser} which will make
this combinator compositional (as opposed to \parser{some} defined in
Section~\ref{parser:some}). This as a cost in terms of the readability of
the definition of \AF{schainl}. But ultimately all of this complexity only
shows up in the implementation of our library: the end user can blissfully
ignore these details.

@@

As we have mentioned when defining \AF{schainl}, the combinator \AF{hchainl}
we have just implemented does not expect fully-defined parsers as arguments.
As a consequence it can be used inside a fixpoint construction. Both the parser
for the constructor and the one for its \AB{B} argument are guarded whilst the
one for the \AB{A} seed is not. This means that trying to define a left-recursive
grammar by immediately using a recursive substructure on the left is now a \emph{type
error}. But it still possible to have some on the right or after having consumed
at least one character (typically an opening parenthesis, cf. the \type{Expr}
example in Section~\ref{parser:expr}).

\section{Fully Worked-Out Example}\label{example:expr}

From \AF{hchainl}, one can derive \AF{chainl1} which is not heterogeneous
and uses the same parser for the seed and the constructors' arguments.
This combinator together with the idea of precedence mentioned in Section~\ref{parser:expr}
is typically used to implement left-recursive grammars. Looking up the
documentation of the \texttt{parsec} library on hackage~\cite{doc:parsec}
we can find a fine example: an extension of our early arithmetic language
(corresponding grammar on the right hand side).

\noindent
\begin{minipage}{0.45\textwidth}
  \shgrab{example}
\end{minipage}
\begin{minipage}{0.55\textwidth}
  \begin{tabular}{lllll}
    Expr   & := & Term   & | & Term AddOp Expr \\
    Term   & := & Factor & | & Factor MulOp Term \\
    Factor & := & <Int>  & | & '(' Expr ')' \\
    AddOp  & := & '+' & | & '-' \\
    MulOp  & := & '*' & | & '/' \\
  \end{tabular}
\end{minipage}

@@

Although quite close to the Haskell version, We can notice four minor changes:

Firstly, the intermediate parsers need to be declared before being used which effectively
reverses the order in which they are spelt out.

Secondly, the recursive calls
are now explicit: in the definition of \type{factor}, \AF{expr} is \AF{map}ped
under the \BOX{} to project the recursive call to the \AD{Expr} \AR{Parser} out
of \AR{Language}.

Thirdly, we use \AF{hchainl} instead of \AF{chainl1} because
breaking the grammar into three distinct categories leads us to parsing
\emph{h}eterogeneous left chains.

Fourthly, we have to insert calls to \AF{box}
to lift \AR{Parser}s into boxed ones whenever the added guarantee that the call
will be guarded is of no use to us. This last point however does not stand in
Coq nor Idris which have a mechanism to declare implicit coercions and where the
typechecker can insert these calls to \AF{box} automatically for us.

@@

Secondly, there is no reason to limit ourselves to \type{Char} as the
unit of information to be processed. Near all of our combinators are
fully polymorphic over the kind of tokens they can deal with. To run
the parser, the user will have to define an appropriate tokenizer for
their use case. The library provides a trivial one for \AD{Char}.

Thirdly, there is no reason to force the user to get back a \AD{List}
of successes: any Functor which is both a Monad and an Alternative
will do. This means in particular that a user may for instance instrument
a parser with a logging ability to be able to return good error messages,
have a (re)configurable grammar using a \type{Reader} transformer or use
a \type{Maybe} type if they want to make explicit the fact that their
grammar is unambiguous.

@@

The reified approach allows him to define a grammar's semantics in terms
of multisets of words and prove sound a variant of Brzozowski derivatives~\citeyear{brzozowski1964derivatives}
as well as study the equational theory of parsers. The current
implementation, based on the Brzozowski derivatives, is however of complexity
at least exponential in the size of the input.

Our approach, although not able to tackle certification like Danielsson's,
is however more lightweight. Using only strong induction on the natural
numbers, it is compatible with languages a lot less powerful than Agda.
Indeed there is no need for good support for mixed induction and coinduction
in the host language. And although we do rely on enforcing invariants
at the type-level, one could mimic these in languages with even weaker type
systems by defining an \emph{abstract} \BOX{} and only providing the user with
our set of combinators which is guaranteed to be safe.
